---
title: "Networks From Retina Part 2"
author: "Lisa"
date: "7/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# Aim

With this analysis we try to answer the following questions; <br>
* Can we reach a sft model fit after data imputation? <br>
* How *good* do the dendrograms look?
* How similiar are the networks?

# Set-up

## Directories

```{r directories}
dir_name <- "/data/bioinf/projects/data/2021_NetworksRetina/"
input <- paste0(dir_name, "input/")
rdata <- paste0(dir_name, "RData/")
results <- paste0(dir_name, "results/")
csvs <- paste0(dir_name, "csvs/")

```

## Packages

```{r packages, error=FALSE, warning=FALSE, message=FALSE}
library("WGCNA")
library("Seurat")
library("ggplot2")
library("forcats")
library("tidyverse")
library(biomaRt)
library("jaccard")
library(reshape2)
library(magrittr)
library(DT)
library(xtable)
```

```{r functions,  error=FALSE, warning=FALSE, message=FALSE}


ret_dca <- readRDS(paste0(rdata,"01_SeuratData_DCA.rds"))
load(paste0(rdata,"01_Subset_Cones.RData"))
load(paste0(rdata,"01_Subset_Rods.RData"))
load(paste0(rdata,"04_MM_cones_DCA.RData"))
load(paste0(rdata,"04_MM_rods_DCA.RData"))


get_top_hubs <- function(datKME_df, n_hubs){
  hub_genes <- list()
  for (i in 1:dim(datKME_df)[2]){
    hub_genes[[i]] <- datKME_df %>% 
      rownames_to_column() %>% 
      arrange(desc(datKME_df[,i])) %>%
      dplyr::select(rowname) %>% 
      dplyr::slice(1:n_hubs)
    hub_genes[[i]] <- hub_genes[[i]]$rowname}
  names(hub_genes) <- gsub(colnames(datKME_df), pattern = "MM.", replacement = "")
  return(hub_genes)
}



ret_dca <- readRDS(paste0(rdata,"01_SeuratData_DCA.rds"))
load(paste0(rdata,"01_Subset_Cones.RData"))
load(paste0(rdata,"01_Subset_Rods.RData"))
load(paste0(rdata,"04_MM_cones_DCA.RData"))
load(paste0(rdata,"04_MM_rods_DCA.RData"))
#' Export the subnetwork densely connected to a particular node to Cytoscape
#' 
#' This function selects a set of nodes based on connecivity to a given node and produced export objects for Cytoscape.  
#' 
#' requires nothing so far 
#' 
#'   
#' @param IDs vector of IDs/gene names matching the rows and columns of the adjacency matrix
#' @param adjacency adjancency matrix
#' @param target ID of the target node
#' @param adj.CO adjacency cutoff
#' @param t.module target module, only nodes within this moule are considered if set
#' @param modules assigned modules to nodes mandatory of t.module is set
#' @param annot annotation data frame that genes in rows annotations in columns
#' @param annot.in.node.t column names of annot columns that should end up in the Cytoscape node table file
#' @param annot.in.edge.t column names of annot columns that should end up in the Cytoscape edge table file
#' @param annot.id.match.col column name of annot column to be used for annotation matching, defalts to 'gene_id'
#' @param base.name file  base name for edge and node table files
#' @param primary.name.col column used populating the name column if this is NA nodeName is used
#' @param kME Matrix of Eigengene correlations for all genes rownames need to correspond to IDs, colnames to modules. If set includes Eigenegenes with abs kME > edj.CO 
#' @param exon.count.CO cutoff from max exons gene, if provided nodes with less exons are discarded
#' @param ignore.exon.xount.protein.coding ignore exon count cutoff for protein coding genes requires annot to contain a column gene_biotype
#' @param edge.weight.transform.exp exponent for transforming edge weight, sign is maintained 
#' @param pos.adj.only only report interactions where the adjacency >= adj.CO, if false abs (adjacency) > = adj.CO nodes and edges are returned, useful when adjacency is actually a signed measure like correlation 
#' @examples 
#' subNet2Cytoscape (target='segemehl_XLOC_068379',t.module='thistle', IDs=colnames (datExpr),adjacency=adjacency,
#' adj.CO=0.7, modules=mergedColors, annot=annot, annot.in.node.t=c('gene_biotype','max_exons_gene', 'AID'), 
#' annot.in.col.t=c(), annot.id.match.col='gene_id', primary.name.col='hgnc_symbol', kME=kmE.signed
#' base.name='WGCNA-08b-XLOC_089379')
#' 
#' 
subNet2Cytoscape  <- function (target,t.module=NULL, IDs,adjacency,adj.CO, exon.count.CO=NULL, modules, annot, annot.in.node.t, annot.in.col.t, 
                               annot.id.match.col='gene_id', primary.name.col='hgnc_symbol', base.name, kME=NULL, ignore.exon.count.protein.coding=F,
                               edge.weight.transform.exp=5, pos.adj.only=T, debug=0, fixed_n = 0) {
  if (!is.null (t.module) && is.null (modules)) {
    stop ('Modules must not be NULL if t.module is given')
  }
  
  target.idx <- match (target, IDs)
  
  ## Selection based on adj.CO
  if (fixed_n != 0){
    diag(adjacency) <- NA
    sel.node <- adjacency [target.idx, ]
    sel.node<- sort(sel.node, decreasing = T)[1:fixed_n]
    # sel.node <- as.logical(sel.node, target.idx)
    sel.node.idx <-as.logical(rownames(adjacency) %in% names(sel.node)+ rownames(adjacency) %in% target)
  } else {
    if (pos.adj.only==T) {
      sel.node.idx <- adjacency [target.idx, ]>= adj.CO
    }
    else {
      sel.node.idx <- abs(adjacency [target.idx, ]) >= adj.CO   
    }}
  
  ## Selection based on module memebership
  if (!is.null (t.module)) {
    sel.node.idx <- sel.node.idx & modules %in% c(t.module)
  }
  
  ## Selection based on exon count
  if (!is.null (exon.count.CO)) {
    if (is.null (annot) | ! 'max_exons_gene' %in% colnames (annot)) {
      stop('If exon.count.CO is set, annot must be provided and contain a max_exons_gene column!')
    }
    if (ignore.exon.count.protein.coding==T) {
      if ('gene_biotype' %in% colnames (annot)) {
        sel.node.idx <- sel.node.idx & (annot[match (IDs, annot[,annot.id.match.col]),'max_exons_gene'] > exon.count.CO | 
                                          annot[match (IDs, annot[,annot.id.match.col]),'gene_biotype']=='protein_coding')
      }
      else {
        stop ('annot needs to contain a column gene_biotype to ignore exon count cutoff for protein coding genes!')
      }
    }
    else {
      sel.node.idx <- sel.node.idx & annot[match (IDs, annot[,annot.id.match.col]),'max_exons_gene'] > exon.count.CO
    }
  }
  
  sel.node.idx <- which (sel.node.idx)
  
  node.table=data.frame (nodeName=IDs[sel.node.idx])
  ## columns to add to node table
  if (!is.null (annot.in.node.t)) {
    cols <- match (annot.in.node.t, colnames (annot))
    if (length (which (is.na(cols))) > 0) {
      stop (paste ('annot.in.node.t argument(s) ', which (is.na(cols)), ' do not match colnames in annot table!', sep="", collapse="," ))
    }
    #cols <- which (cols)
  }
  if (!is.null (annot)) {
    if (length (which (is.na(match (annot.id.match.col, colnames (annot))))) > 0) {
      stop (paste ('annot.id.match.col=', annot.id.match.col, ' does not match any column name in annot!', sep=''))
    }
    node.table  <- cbind (node.table, annot [match (node.table$nodeName, annot[,annot.id.match.col]), cols])
  }
  ## Generate a name column that contains either the symbol 
  if (debug > 0) message ('Constructing node table')
  node.table$pname  <- annot [match (node.table$nodeName, annot[,annot.id.match.col]),primary.name.col]
  node.table$pname[is.na(node.table$pname) | is.null (node.table$pname) | node.table$pname==""] <- 
    node.table$nodeName[is.na(node.table$pname) | is.null (node.table$pname) | node.table$pname==""]
  ## Add moduel informatoin if available
  if (!is.null(modules)) {
    node.table$module <- modules[sel.node.idx]
  }
  ## Add a column type to discriminate genes from modules 
  node.table$type='gene'
  
  ## Make some corrections to NA entries in tables
  if ('gene_biotype' %in% colnames (node.table)) {
    node.table$gene_biotype [is.na(node.table$gene_biotype)]  <- 'XLOC'
  }
  
  ## if kME is provided, add eigengene correlation for the selected nodes 
  if (! is.null (kME)) {
    tmp.table  <- matrix (nrow=dim (kME)[2], ncol=dim (node.table)[2], data=NA)
    #message (paste ('dim tmp table:', unlist(dim(tmp.table)),sep="", collapse=","))
    colnames (tmp.table) <- colnames (node.table)
    tmp.table <- as.data.frame (tmp.table)
    tmp.table$nodeName=colnames(kME)
    tmp.table$pname <- colnames(kME)
    tmp.table$type='module'
    tmp.table  <- tmp.table [apply (kME[sel.node.idx,], 2, function(x) {sum (as.numeric (abs(x)> adj.CO))>0}),]
    node.table=rbind (node.table, tmp.table)
  }
  if (debug > 0) message ('Finished node table')
  
  ### Edge table
  if (debug > 0) message ('Constructing edge table')
  
  
  
  edge.table  <-  melt (adjacency[sel.node.idx, sel.node.idx], value.name='adjacency',  id.vars=c())
  colnames (edge.table)=c('fromNode', 'toNode','adjacency')
  ## Remove former diagonal elements
  message ('Removing diagnonal elements from edge table')
  edge.table <- edge.table[edge.table$fromNode!=edge.table$toNode,]
  
  
  ## Remove adges with adjacency below cutoff
  if (pos.adj.only==T) {
    edge.table  <- edge.table[edge.table$adjacency >= adj.CO,]
  }
  else {
    edge.table  <- edge.table[abs(edge.table$adjacency) >= adj.CO,] 
  }
  
  # if(fixed_n != 0){
  #   edge.table <- edge.table %>% arrange(desc(adjacency)) %>%
  #     dplyr::slice(1:fixed_n)
  # }
  ## Add edges to Eigengenes, if kME has been provided
  if (! is.null (kME)) {
    edge.tmp  <- melt (cbind (row.names(kME[sel.node.idx, ]), kME[sel.node.idx,]), value.name='adjacency', id.vars=c(1))
    colnames (edge.tmp)=c('fromNode', 'toNode','adjacency')
    edge.tmp <- edge.tmp[abs(edge.tmp$adjacency) >= adj.CO,]
    edge.table <- rbind (edge.table, edge.tmp)
  }
  ## Transform edge weight and provide unsigned weight
  edge.table$abs.adjacency <- abs(edge.table$adjacency)
  edge.table$transf.adjacency  <- sign (edge.table$adjacency) * abs (edge.table$adjacency)^edge.weight.transform.exp
  edge.table$abs.transf.adjacency  <-  abs (edge.table$adjacency)^edge.weight.transform.exp
  message ('Finished edge table')
  ## write node and edge table 
  write.table (node.table, file=paste(csvs, base.name, "-Cytoscape-nodes.csv", sep=""), sep="\t", quote=F,row.names=F)
  write.table (edge.table, file=paste(csvs, base.name , "-Cytoscape-edges.csv",sep=""), sep="\t", quote=F, row.names=F)
  
  print(dim(node.table)[1])
  return(node.table$nodeName)
}

#' Retrieve or load a biomart table containing the selected Gene_attributes for the provided Gene_IDs
#' 
#' @param Gene_IDs        A list of gene IDs for which the biomaRt annotation should be retreived
#' @param ID_type         the type of the Gene_IDs. Default: "ensembl_gene_id"
#' @param Gene_attributes A vector of attributes that should be returned. Note that this function will 
#'                        return a row for the requested gene only if it has a value for ALL attributes!
#'                        You can retrieve a list of attributes via 'listAttributes(ensembl)' (with a loaded mart!)
#' @param biomart         Provide your preferred biomaRt web service. You can get a full list via 'listMarts()'
#' @param dataset         Dataset you want to use. To see the different datasets available within a biomaRt you can 
#'                        e.g. do: mart = useMart('ensembl'), followed by listDatasets(mart).
#' @param host            Host to connect to if different then www.biomart.org 
#' 
get_biomart_table <- function(Gene_IDs, ID_type, Gene_attributes) {
  
  # ids <- gsub("\\.\\d$","", Gene_IDs, perl=TRUE) ## Remove version from gene ids if present
  
  myMart <- useMart("ensembl", dataset="hsapiens_gene_ensembl")
  bmTable <- getBM (attributes = Gene_attributes,
                    filters = ID_type, 
                    values = Gene_IDs, 
                    mart=myMart
  )
  return(bmTable)
}

all_genes <- rownames(cones)


annot <- get_biomart_table(Gene_IDs = rownames(cones), ID_type = "hgnc_symbol", Gene_attributes = c('hgnc_symbol', 'description','gene_biotype'))
save(annot, file = paste0(rdata, '002_Annot_cones_DCA.RData'))

annot <- get_biomart_table(Gene_IDs = rownames(rods), ID_type = "hgnc_symbol", Gene_attributes = c('hgnc_symbol', 'description','gene_biotype'))
save(annot, file = paste0(rdata, '002_Annot_rods_DCA.RData'))


export_genes_cyto <- function(which_dynamic, which_adj, which_annot, target_gene, threshold, which_ntwk, fixed_n){
  load(which_dynamic)
  load(which_adj)
  load(which_annot)
  
  geneList <- subNet2Cytoscape (target=target_gene, IDs=all_genes,adjacency=adjacency_complete,
                                adj.CO=threshold, modules=dynamicColors, annot=annot, annot.in.node.t=c('gene_biotype'), 
                                annot.in.col.t=c(), annot.id.match.col='hgnc_symbol', primary.name.col='hgnc_symbol',
                                base.name=paste0('WGCNA-',which_ntwk,'-',target_gene), fixed_n = fixed_n)
  return(geneList)
  
}



```

## Results


```{r data_prep,error=FALSE, warning=FALSE, message=FALSE}

adjacency <-  WGCNA::adjacency(t(cones), type = "signed" , power = 12, corFnc = "bicor")
adjacency_complete <- adjacency
save(adjacency_complete, file = paste0(rdata, "001_Adj_cones_DCA.RData" ))

load("/data/bioinf/projects/data/2021_NetworksRetina/RData/02_dynamicColors_DCA_Cones.RData")
connecVals_cones <- intramodularConnectivity(adjacency, dynamicColors, scaleByMax = FALSE)

adjacency <-  WGCNA::adjacency(t(rods), type = "signed" , power = 12, corFnc = "bicor")
adjacency_complete <- adjacency
save(adjacency_complete, file = paste0(rdata, "001_Adj_rods_DCA.RData" ))


load("/data/bioinf/projects/data/2021_NetworksRetina/RData/02_dynamicColors_DCA_Rods.RData")
connecVals_rods <- intramodularConnectivity(adjacency, dynamicColors, scaleByMax = FALSE)

```

```{r}


```

```{r calculate, error=FALSE, warning=FALSE, message=FALSE}
hub_gene_df_cones <- get_top_hubs(datKME_df= datKME_cones, n_hubs=30)
hub_gene_df_rods <- get_top_hubs(datKME_df= datKME_rods, n_hubs=30)
all_hubs <- Reduce(intersect, list(unlist(hub_gene_df_cones), unlist(hub_gene_df_rods)))

all_hubs <- all_hubs[all_hubs %in% annot$hgnc_symbol]


summary_cones <- list()
summary_rods <- list()
intersect_vec <- c()
Kcones <- c()
Krods <- c()
Delta_target_abs <- c()
Delta_target_norm <- c()
Delta_whole_network_abs <- c()
Delta_whole_network_norm <- c()
jaccard_vec <- c()
ctrl_jaccard <- c()
for ( i in 1:length(all_hubs)){
  cat(paste("Target_gene:", all_hubs[i], "-----------------------------------------------\n"))
  
  
  cones_gene <-export_genes_cyto(which_dynamic= "/data/bioinf/projects/data/2021_NetworksRetina/RData/02_dynamicColors_DCA_Cones.RData", 
                                 which_adj = paste0(rdata, "001_Adj_cones_DCA.RData"), 
                                 which_annot = paste0(rdata, '002_Annot_cones_DCA.RData'), 
                                 target_gene = all_hubs[i], 
                                 threshold = 0.05, 
                                 which_ntwk = 'cones', 
                                 fixed_n = 300)
  summary_cones[[i]] <- cones_gene
  
  k1 <- connecVals_cones[rownames(connecVals_cones) %in% summary_cones[[i]], "kWithin"]
  Kcones <- c(Kcones, round(mean(k1),2))
  
  # 140
  
  rods_gene <- export_genes_cyto(which_dynamic= "/data/bioinf/projects/data/2021_NetworksRetina/RData/02_dynamicColors_DCA_Rods.RData", 
                                 which_adj = paste0(rdata, "001_Adj_rods_DCA.RData"), 
                                 which_annot = paste0(rdata, '002_Annot_rods_DCA.RData'), 
                                 target_gene = all_hubs[i], 
                                 threshold = 0.05, 
                                 which_ntwk = 'rods', 
                                 fixed_n = 300)
  summary_rods[[i]] <- rods_gene
  
  k2 <- connecVals_rods[rownames(connecVals_rods) %in% summary_rods[[i]], "kWithin"]
  # 136
  
  Krods <- c(Krods, round(mean(k2),2))
  
  Delta_target_abs <- c(Delta_target_abs, abs(round(connecVals_cones[rownames(connecVals_cones) %in% all_hubs,"kWithin"],1)[i] - round(connecVals_rods[rownames(connecVals_rods) %in% all_hubs,"kWithin"],1)[i]))
  Delta_target_norm <- c(Delta_target_norm, abs((round(connecVals_cones[rownames(connecVals_cones) %in% all_hubs,"kWithin"],1)[i]/max(k1)) - (round(connecVals_rods[rownames(connecVals_rods) %in% all_hubs,"kWithin"],1)[i]/max(k2))))
  
  Delta_whole_network_abs <- c(Delta_whole_network_abs, abs(round(mean(k1),2) - round(mean(k2),2)))
  Delta_whole_network_norm <- c(Delta_whole_network_norm, abs(round(mean(k1),2)/max(k1) - round(mean(k2),2)/max(k2)))
  
  cat(paste("Intersect:", length(intersect(cones_gene, rods_gene)), "-----------------------------------------------\n"))
  intersect_vec <- c(intersect_vec, length(intersect(cones_gene, rods_gene)))
  # 35
  
  df <- data.frame(Genes = union(cones_gene, rods_gene)) 
  df <- df %>% 
    mutate(In_cones = ifelse(df$Genes %in% cones_gene, 1, 0)) %>% 
    mutate(In_rods = ifelse(df$Genes %in% rods_gene, 1, 0))
  
  cat(paste("Jaccard:", jaccard(df$In_cones, df$In_rods), "-----------------------------------------------\n"))
  jaccard_vec <- c(jaccard_vec, round(jaccard(df$In_cones, df$In_rods),2))
  cat(paste("Control-Jaccard:", length(intersect(cones_gene, rods_gene))/ length(union(cones_gene, rods_gene)), "-----------------------------------------------\n"))
  ctrl_jaccard <- c(ctrl_jaccard, length(intersect(cones_gene, rods_gene))/ length(union(cones_gene, rods_gene)))
  
}

names(summary_cones) <- all_hubs
names(summary_rods) <- all_hubs
names(Kcones) <- all_hubs
names(Krods) <- all_hubs
names(Delta_target_abs) <- all_hubs
names(Delta_target_norm) <- all_hubs
names(Delta_whole_network_abs) <- all_hubs
names(Delta_whole_network_norm) <- all_hubs
names(intersect_vec) <- all_hubs
names(jaccard_vec)  <- all_hubs
names(ctrl_jaccard)  <- all_hubs

```



```{r datatable}
tissue_info <- read.table('/work/steinheu/tissue_category_rna_retina_Tissue.tsv', sep = '\t', header = T)

df <- annot[annot$hgnc_symbol %in% all_hubs,]
# reorder rows by the order of the all_hubs vector
rownames(df) <- df$hgnc_symbol
df <- df[all_hubs,]
df$'Gene included in human protein atlas list?' <- ifelse(all_hubs %in% tissue_info$Gene, "yes", "no")
df$'K Target Gene in Cones(kWithin)' <- round(connecVals_cones[rownames(connecVals_cones) %in% all_hubs,"kWithin"],1)
df$'K Target Gene in Rods(kWithin)' <- round(connecVals_rods[rownames(connecVals_rods) %in% all_hubs,"kWithin"],1)
df$'Delta K Target (absolute)' <- Delta_target_abs
df$'Delta K Target (norm)' <- Delta_target_norm
df$'K Target NTWK from Cones(Mean(kWithin))' <- Kcones
df$'K Target NTWK from Rods(Mean(kWithin))' <- Krods
df$'Delta K NTWK (absolute)' <-Delta_whole_network_abs
df$'Delta K NTWK (norm)' <-Delta_whole_network_norm
df$'Intersect' <- intersect_vec
df$'Tanimoto Similarity' <- jaccard_vec

datatable(df)


```

```{r save_data, error=FALSE, warning=FALSE, message=FALSE}
save(df, file = paste0(rdata, '001_Final_df_jaccard.RData'))

df %>% dplyr::select(c("hgnc_symbol","K Target Gene in Cones(kWithin)", "K Target Gene in Rods(kWithin)", "Intersect", "Tanimoto Similarity" )) %>% arrange(Intersect) %>% xtable()

melt(summary_cones) %>% set_colnames(c("Ntwk", "Target"))  %>% write_csv(path = paste0(csvs, "Summary_NTWK_cones.csv"))
melt(summary_rods) %>% set_colnames(c("Ntwk", "Target"))  %>% write_csv(path = paste0(csvs, "Summary_NTWK_rods.csv"))

```

Where are the preserved hubs located?
```{r}
# Cones
# black        blue       brown       green greenyellow         red   turquoise      yellow 
#   6           1           1           1          17           5           5           1 
# Rods
# blue     brown     green      pink    purple       red turquoise    yellow 
#  17         7         1         1         5         1         4         1 

```

It appear that most from teh cone ntwk are located in the greenyellow module and from the Rod ntwk in the blue module.


# EIF4EBP1
Min jaccard
```{r}
cones_gene <-export_genes_cyto(which_dynamic= "/work/steinheu/2021_NetworksFromRetina/RData/02_dynamicColors_DCA_Cones.RData", 
                               which_adj = paste0(rdata, "001_Adj_cones_DCA.RData"), 
                               which_annot = paste0(rdata, '002_Annot_cones_DCA.RData'), 
                               target_gene = 'EIF4EBP1',
                               threshold = 0.05, 
                               which_ntwk = 'cones', 
                               fixed_n = 30)

rods_gene <- export_genes_cyto(which_dynamic= "/work/steinheu/2021_NetworksFromRetina/RData/02_dynamicColors_DCA_Rods.RData", 
                               which_adj = paste0(rdata, "001_Adj_rods_DCA.RData"), 
                               which_annot = paste0(rdata, '002_Annot_rods_DCA.RData'), 
                               target_gene = 'EIF4EBP1',
                               threshold = 0.05, 
                               which_ntwk = 'rods', 
                               fixed_n = 30)

cat(paste("Intersect between both small networks:", length(intersect(cones_gene, rods_gene)),"---------------------------------"))
```


