---
title: "Untitled"
author: "Lisa"
date: "October 14, 2020"
output: html_document
-
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Information
- Paper : https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6596422/
- Human peripheral and foveal cells sequenced

# Data set up 
```{r}

input <- "/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/input/"
rdata <- "/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/RData/"
results <- "/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/results/"
project <- "Roska_Retina"

# The data was retrieved from: https://data.iob.ch/
```

# Functions

```{r}
library("tidyverse")
library("rhdf5")
library("plyr")
library('data.frame')

theme_basic2 <- function() {
  theme_bw(base_size = 15) +
    theme(strip.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank())
}

PreprocessRetinaData <- function(data_name){
  setwd(input)
  mat <- fread(paste0(data_name,"_exprMatrix.tsv"), header = T)
  meta <- read.table(paste0(data_name,"_meta.tsv"), header=T, sep="\t", as.is=T, row.names=1)
  genes = mat[,1][[1]]
  genes = gsub(".+[|]", "", genes)
  mat = data.frame(mat[,-1], row.names=genes)
  colnames(mat) <- rownames(meta)
  annot <- as.data.frame(t(mat))
  rm(mat)
  annot$CellType <- meta$cell_type
  annot$Donor <- meta$subject
  for (i in unique(meta$subject)){
    annot_sub <- annot %>% filter(Donor == i) %>% select(-Donor)
    means_Rods_data <<- SubsampleCelltype(data_input = annot_sub,
                                          celltype = 'rod')
    write.table(means_Rods_data, file = paste0(rdata, "01_", project, "_MeansRods_", data_name, "_", i ,".tsv"), sep = "\t", dec = ",")
    means_Cones_data <<- SubsampleCelltype(data_input = annot_sub,
                                           celltype = 'cone')
    write.table(means_Cones_data, file = paste0(rdata, "01_", project, "_MeansCones_", data_name,"_", i , ".tsv"), sep = "\t", dec = ",")
    means_MG_data <<- SubsampleCelltype(data_input = annot_sub,
                                        celltype = 'MG')
    write.table(means_MG_data, file = paste0(rdata, "01_", project, "_MeansMuller-Glia_", data_name,"_", i , ".tsv"), sep = "\t" , dec = ",")
  }
  rm(annot)
}

SubsampleCelltype <- function(data_input, celltype){
  if (celltype == 'cone'){
    Subsampled_df <- data_input %>%
      filter(CellType == "L/M cone" | CellType == "S cone") %>%
      select(-CellType)
  } else if (celltype == 'MG'){
    Subsampled_df <- data_input %>%
      filter(CellType == "MC_01") %>%
      select(-CellType)
  } else {Subsampled_df <- data_input %>%
    filter(CellType == celltype) %>%
    select(-CellType)}
  means_Subsampled_df <<- data.frame("Genes" = colnames(Subsampled_df),
                                     "GeneMean" = colMeans(Subsampled_df))
}

Calculate_correlation <- function(data_type, intersect, sc_data, pattern_data){
  # Preparation of data
  files <- list.files(path=rdata, pattern = pattern_data)
  for(i in 1:length(files)){
    means_data <- read.table(paste0(rdata,files[i]), sep = "\t", row.names = 1, dec = ",")
    means_filtered <- means_data[intersect,]
    if (i ==1){
      combined <- means_filtered
    } else {
      combined <- cbind(combined, means_filtered$GeneMean)
    }
  }
  combined <- combined[-1]
  colnames(combined) <- gsub(files, pattern ="0[0-9]_Roska_Retina_Means|.tsv", replacement = "")
  print( colnames(combined))
  combined_sc <<- cbind(combined, t(sc_data[,intersect]))
  print("Data assembled!")
  save(combined_sc, file = paste0(rdata, "03_", project, "_CombinedData", data_type,".RData"))
  print("Starting correlation calculation")
  # Calculate Correlation
  correlation_mat <<- cor(combined_sc)
  save(correlation_mat , file = paste0(rdata, "01_Correlation_", data_type,".RData"))
}

# Plot_data <- function(data_type, correlation_df, data_included){
#   if (data_included == "MG"){
#     svg(paste0(results,"01_",project, "_Correlation_MG_",data_type ,".svg"),
#         width = 15, height = 8)
#     p <<-  ggplot(correlation_df %>% filter(Celltype  == "MuellerGlia"),
#                   aes(x = Retina.type, y = Correlation, group = Retina.type, fill = Groups)) +
#       geom_violin(trim=FALSE) +
#       geom_boxplot(width=0.1) +
#       xlab("Retina type") +
#       ylab("Correlation") +
#       theme_basic2() +
#       theme(axis.text.x=element_blank(),
#             axis.ticks.x=element_blank()) +
#       labs(title = "Correlation towards peripheral and foveal retina \n in Mueller Glial cells",
#            subtitle = paste(data_type," - Annotation based on higher resolution"),
#            caption = "Mean expression value from Glia cells separately \n
#                   from foveal and peripheral samples \n
#                   One group includes samples from three donors") +
#       scale_fill_manual(values=c("#0000FF", "#6363F3")) 
#       #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
#       #facet_grid(cols = vars(Celltype))
#     print(p)
#     dev.off()
#   } else if (data_included == "R/C") {
#     svg(paste0(results,"01_",project, "_Correlation_Cones_Rods_",data_type ,".svg"),
#         width = 15, height = 8)
#     p <<-  ggplot(correlation_df %>% filter(Celltype  == c("Rods", "Cones")),
#                   aes(x = Retina.type, y = Correlation, group = Retina.type, fill = Groups)) +
#       geom_violin(trim=FALSE) +
#       geom_boxplot(width=0.1) +
#       xlab("Retina type") +
#       ylab("Correlation") +
#       theme_basic2() +
#       theme(axis.text.x=element_blank(),
#             axis.ticks.x=element_blank()) +
#       labs(title = "Correlation towards peripheral and foveal retina \n in Rods and Cones",
#            subtitle = paste(data_type," - Annotation based on higher resolution"),
#            caption = "Mean expression value from cones and rods separately \n
#                   from foveal and peripheral samples \n
#                   One group includes samples from three donors") +
#       scale_fill_manual(values=c("#BA2BB2", "#F79CF2", "#0BA8AF", "#9BF4F8")) +
#       #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
#       facet_grid(cols = vars(Celltype))
#     print(p)
#     dev.off()}
# }

# Which cluster corresponds to which cell type --> found in Figure 1 
# Information stored in "cluster_label"
# Rods <- c(1,2)
# Cones <- c(3,4)
# MG <- c(13:17)
```

# Read in data

# Analysis
## schROHT Control 2 sample

```{r}
# library("rhdf5")
# 
# h5ad_file <- paste0(input,"20200713_CTRL_2_Cells_higherResolution_pp.h5ad")
# 
# obs <- h5read(h5ad_file, name = "obs")

load('/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/RData/CaSTLe_Annot_output_Control_2_Organoid_cellnames.RData')

sc_data <- read.csv("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/csvs/Count_data_schROHT_Control2.csv", row.names = 1, header = T)
sc_data_log<- log10(sc_data+1)
```

### Subsampling Scheetz data

```{r}
all_files <- c("Foveal", "Peripheral")
for (i in 1:length(all_files)){
  PreprocessRetinaData(data_name = all_files[i])
}

#intersect_genes <- Reduce(intersect,  L_genes)
intersect_full <- intersect(colnames(sc_data), means_Subsampled_df$Genes)

```


### Rods and Cones
#### Correlation

```{r}
Calculate_correlation(data_type = "Control2", intersect = intersect_full, sc_data = sc_data_log, pattern_data = "01_Roska_Retina_*")


```

### Annotate Louvain clusters according to Master GOI expression

```{r}
load(paste0(rdata, "01_Correlation_Control2.RData"))
red_mat <- correlation_mat[-c(1:18), 1:18]
# rownames(obs) <- obs$index
cor_louvain <- cbind(red_mat, predictedClasses)
colnames(cor_louvain) <- c(colnames(red_mat), "Louvain")

cor_louv <- data.frame("Retina type" = as.factor(c(rep(colnames(cor_louvain)[-dim(cor_louvain)[2]], each = dim(cor_louvain)[1]))),
                       "Correlation" = as.vector(red_mat),
                       "Celltype"= as.factor(rep(cor_louvain[,dim(cor_louvain)[2]],dim(red_mat)[2])))


cor_louv$Celltype <- revalue(cor_louv$Celltype, replace = c("MC" = "Muller-Glia",
                                                            "cones" = "Cones",
                             "rod" = "Rods"))
cor_louv$Groups <- gsub(cor_louv$Retina.type, pattern = "_R-00646_0[1-4]", replacement = "")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = "_", replacement = " ")
cor_louv$Correspondance <- ifelse((gsub(cor_louv$Groups, pattern = " Foveal", replacement = "") == cor_louv$Celltype | 
                                     gsub(cor_louv$Groups, pattern = " Peripheral", replacement = "") == cor_louv$Celltype) , TRUE,FALSE)

cor_louv$Groups <- gsub(cor_louv$Groups, pattern = "Muller-Glia", replacement = "Müller Glia")
cor_louv$Celltype <- gsub(cor_louv$Celltype, pattern = "Muller-Glia", replacement = "Müller Glia")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = " Foveal", replacement = " fovea")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = " Peripheral", replacement = " peripheral")
cor_louv$Cellname <- rep(rownames(cor_louvain), 18)
```
```{r}
# Heatmap of peripheral and foveal vectors
pheatmap(correlation_mat[1:12,1:12], col = brewer.pal(9,"Reds"), cluster_rows=FALSE, cluster_cols=FALSE)
```

### Plot

```{r}
svg(paste0(results,"01_",project, "_Alt_Figure_Correlation_Control2_CaSTLE.svg"),
    width = 10, height = 5)
ggplot(
  cor_louv %>% filter(Celltype  == "Rods"| Celltype  =="Cones" | Celltype  =="Müller Glia") %>%
    filter(Correspondance == TRUE),
  aes(x = Groups, y = Correlation, fill= Groups)) +
  geom_violin(trim=FALSE) +
  geom_boxplot(width = 0.1) +
  coord_cartesian(ylim = c(0,1)) +
  xlab("Retina type") +
  ylab("Correlation") +
  theme_basic2() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank(),
        text=element_text(size=20),
        legend.position="bottom") +
  # labs(title = "Correlation of schROHT photoreceptors and MG cells \n towards peripheral and foveal reference data",
  #      subtitle = paste("Control 2"," - Annotation based on higher resolution"),
  #      caption = expression(paste("Reference data taken from Voigt ", italic("et. al")))) +
  scale_fill_manual(values=c("#BA2BB2", "#F79CF2","#0000FF", "#6363F3", "#0BA8AF", "#9BF4F8")) +
  #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
  facet_wrap(vars(Celltype), scales = "free_x")
dev.off()

svg(paste0(results,"01_",project, "_AllDonors_Correlation_Control2_CaSTLe.svg"),
    width = 10, height = 12)
ggplot(
  cor_louv %>% filter(Celltype  == "Rods"| Celltype  =="Cones" | Celltype  =="Müller Glia") %>%
    filter(Correspondance == TRUE),
  aes(x = Retina.type, y = Correlation, fill= Groups)) +
  geom_violin(trim=FALSE) +
  geom_boxplot(width = 0.1) +
  coord_cartesian(ylim = c(0,1))+
  xlab("Retina type") +
  ylab("Correlation") +
  theme_basic2() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank(),
        text=element_text(size=20),
        legend.position="bottom")+
  # labs(title = "Correlation of schROHT photoreceptors and MG cells \n towards peripheral and foveal reference data",
  #      subtitle = paste("Control 2"," - Annotation based on higher resolution"),
  #      caption = expression(paste("Reference data taken from Voigt ", italic("et. al")))) +
  scale_fill_manual(values=c("#BA2BB2", "#F79CF2","#0000FF", "#6363F3", "#0BA8AF", "#9BF4F8")) +
  #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
  facet_wrap(vars(Celltype), scales = "free_x", nrow = 3)

dev.off()

# Plot_data(data_type = "Control2", correlation_df = cor_louv)

```

```{r}
library("rhdf5")

# h5ad_file <- paste0(input,"20200713_CTRL_3_Cells_higherResolution_pp.h5ad")
# 
# obs <- h5read(h5ad_file, name = "obs")

load('/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/RData/CaSTLe_Annot_output_Control_3_Organoid_cellnames.RData')

sc_data <- read.csv("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/csvs/Count_data_schROHT_Control3.csv", row.names = 1, header = T)
sc_data_log<- log10(sc_data+1)
```

### Subsampling Scheetz data

```{r}
all_files <- c("Foveal", "Peripheral")
for (i in 1:length(all_files)){
  PreprocessRetinaData(data_name = all_files[i])
}

#intersect_genes <- Reduce(intersect,  L_genes)
intersect_full <- intersect(colnames(sc_data), means_Subsampled_df$Genes)

```


### Rods and Cones
#### Correlation

```{r}
Calculate_correlation(data_type = "Control3", intersect = intersect_full, sc_data = sc_data_log, pattern_data = "01_Roska_Retina_*")


```

### Annotate Louvain clusters according to Master GOI expression

```{r}
load(paste0(rdata, "01_Correlation_Control3.RData"))
red_mat <- correlation_mat[-c(1:18), 1:18]
# rownames(obs) <- obs$index
cor_louvain <- cbind(red_mat, predictedClasses)
colnames(cor_louvain) <- c(colnames(red_mat), "Louvain")

cor_louv <- data.frame("Retina type" = as.factor(c(rep(colnames(cor_louvain)[-dim(cor_louvain)[2]], each = dim(cor_louvain)[1]))),
                       "Correlation" = as.vector(red_mat),
                       "Celltype"= as.factor(rep(cor_louvain[,dim(cor_louvain)[2]],dim(red_mat)[2])))

cor_louv$Celltype <- revalue(cor_louv$Celltype, replace = c("MC" = "Muller-Glia",
                                                            "cones" = "Cones",
                             "rod" = "Rods"))
cor_louv$Groups <- gsub(cor_louv$Retina.type, pattern = "_R-00646_0[1-4]", replacement = "")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = "_", replacement = " ")
cor_louv$Correspondance <- ifelse((gsub(cor_louv$Groups, pattern = " Foveal", replacement = "") == cor_louv$Celltype | 
                                     gsub(cor_louv$Groups, pattern = " Peripheral", replacement = "") == cor_louv$Celltype) , TRUE,FALSE)

cor_louv$Groups <- gsub(cor_louv$Groups, pattern = "Muller-Glia", replacement = "Müller Glia")
cor_louv$Celltype <- gsub(cor_louv$Celltype, pattern = "Muller-Glia", replacement = "Müller Glia")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = " Foveal", replacement = " fovea")
cor_louv$Groups <- gsub(cor_louv$Groups, pattern = " Peripheral", replacement = " peripheral")
cor_louv$Cellname <- rep(rownames(cor_louvain), 18)
```
```{r}
# Heatmap of peripheral and foveal vectors
pheatmap(correlation_mat[1:12,1:12], col = brewer.pal(9,"Reds"), cluster_rows=FALSE, cluster_cols=FALSE)
```

### Plot

```{r}
svg(paste0(results,"01_",project, "_Alt_Figure_Correlation_Control3_CaSTLE.svg"),
    width = 10, height = 5)
ggplot(
  cor_louv %>% filter(Celltype  == "Rods"| Celltype  =="Cones" | Celltype  =="Müller Glia") %>%
    filter(Correspondance == TRUE),
  aes(x = Groups, y = Correlation, fill= Groups)) +
  geom_violin(trim=FALSE) +
  geom_boxplot(width = 0.1) +
  coord_cartesian(ylim = c(0,1)) +
  xlab("Retina type") +
  ylab("Correlation") +
  theme_basic2() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank(),
        text=element_text(size=20),
        legend.position="bottom") +
  # labs(title = "Correlation of schROHT photoreceptors and MG cells \n towards peripheral and foveal reference data",
  #      subtitle = paste("Control 2"," - Annotation based on higher resolution"),
  #      caption = expression(paste("Reference data taken from Voigt ", italic("et. al")))) +
  scale_fill_manual(values=c("#BA2BB2", "#F79CF2","#0000FF", "#6363F3", "#0BA8AF", "#9BF4F8")) +
  #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
  facet_wrap(vars(Celltype), scales = "free_x")
dev.off()

svg(paste0(results,"01_",project, "_AllDonors_Correlation_Control3_CaSTLe.svg"),
    width = 10, height = 12)
ggplot(
  cor_louv %>% filter(Celltype  == "Rods"| Celltype  =="Cones" | Celltype  =="Müller Glia") %>%
    filter(Correspondance == TRUE),
  aes(x = Retina.type, y = Correlation, fill= Groups)) +
  geom_violin(trim=FALSE) +
  geom_boxplot(width = 0.1) +
  coord_cartesian(ylim = c(0,1))+
  xlab("Retina type") +
  ylab("Correlation") +
  theme_basic2() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.x=element_blank(),
        text=element_text(size=20),
        legend.position="bottom")+
  # labs(title = "Correlation of schROHT photoreceptors and MG cells \n towards peripheral and foveal reference data",
  #      subtitle = paste("Control 2"," - Annotation based on higher resolution"),
  #      caption = expression(paste("Reference data taken from Voigt ", italic("et. al")))) +
  scale_fill_manual(values=c("#BA2BB2", "#F79CF2","#0000FF", "#6363F3", "#0BA8AF", "#9BF4F8")) +
  #scale_color_manual(values=c("#BA2BB2", "#EB59E3", "#13BBC3", "#2EE6EE")) +
  facet_wrap(vars(Celltype), scales = "free_x", nrow = 3)

dev.off()

# Plot_data(data_type = "Control2", correlation_df = cor_louv)

```
