---
title: "CaSTLe transfer learning schROHT"
author: "Lisa"
date: "May 27, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, message=F}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("SingleCellExperiment")
# 
# 
# BiocManager::install("scater")
# 

library(SingleCellExperiment)
library(scater)  # tested on version 1.6.3,    install from Bioconductor: source("https://bioconductor.org/biocLite.R"); biocLite("scater")
library(xgboost) # tested on version 0.6.4.1, install from CRAN: install.packages("xgboost")
library(igraph)  # tested on version 1.2.1,   install from CRAN: install.packages("igraph")
library(caret)
library(e1071)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(reshape2)
```

# Parameters and directories

```{r parameters_functions}

# Parameters used for CaSTLe run
#-----------------------------------------------------------------------------------------#

BREAKS=c(-1, 0, 1, 6, Inf)
nFeatures = 100


# Functions
#-----------------------------------------------------------------------------------------#

#' Visualize the training stats from the xgboost training section
#' Plots a line plot showing Sensitivity and Specificity
#' over every cell type from Cowan paper.
#' The plot is sorted by Sensitivity and Specificity
#'
#' @param input_df dataframe of melted confusion matrix
#' @param raw_df dataframe where [,1] is true label and [,2] is predicted label
#' @param which_sample chr, name of data used for plotting and saving
#'
#' @return ggplot object
#' 
#' @examples visual_Training_stats(input_df = df, 
#'                            raw_df = eval_df, 
#'                            which_sample = "Control2")

visual_Training_stats <- function(input_df, raw_df, which_sample){
  p <- input_df %>%
    mutate(Celltype = fct_reorder(Celltype, value)) %>%
    ggplot(aes(x = Celltype, y = value, colour = Measure, group = Measure)) +
    geom_point(size = 1.5) +
    geom_line(size = 1.5) +
    theme_bw(base_size = 26) +
    coord_cartesian(ylim = c(0,1)) +
    scale_y_continuous(breaks=c(0,0.5,1)) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1),
          legend.position="bottom") +
    ylab("")
    # Plot labeling
    # labs(title = paste("Evaluating the xgboost from HRO", which_sample),
    #      subtitle = "20% held out data from Cowan paper",
    #      caption = paste(dim(raw_df)[1],"cells"))
  
  ggsave(file = paste0(results,"R_01_Eval_xgboost_",which_sample,".pdf"), 
         width = 10, 
         height= 6,
         plot = p, 
         device = "pdf", 
         limitsize = FALSE)
  
  return(p)
}


analyze_CaSTLe_run <- function(eval_data, cellname_vec, annot_data, data_type, control_data){
  # 9.2 Evaluate classifyer 
  load(eval_data)
  x <- load(cellname_vec)
  new_celltypes <- get(x)
  
  conf_data <- confusionMatrix(data = as.factor(eval_df$pred_labels), 
                               reference = as.factor(eval_df$true_labels))
  # 
  # if(length(as.character(levels(new_celltypes)))- length(rownames(conf_data$byClass)) == 0){
  dic <- as.character(levels(new_celltypes)[c(as.numeric(gsub(rownames(conf_data$byClass), pattern = "Class: ", replacement = ""))+1)])
  rownames(conf_data$byClass) <- dic
  # print(conf_data$byClass)
  # } else {
  #   print("Some cell type was not classifiable! Please check!")
  #   break
  # }
  
  df <- melt(conf_data$byClass[,1:2]) %>%
    set_colnames(c("Celltype", "Measure","value"))
  
  
  p <- visual_Training_stats(input_df = df, 
                             raw_df = eval_df, 
                             which_sample = paste0(control_data, "_", data_type))
  
  print(p)
  
  # Load predicted vector
  load(annot_data)
  
  # Create the 'dictionary' with the initial facor levels
  dic <- as.character(levels(new_celltypes))
  
  # For each entry in the predicted Vector, take the value
  # plus one and look in the dictionary and replace the value
  for(i in 1:length(predictedClasses)) {
    predictedClasses[i] <- dic[as.numeric(predictedClasses[i])+1]}
  
  print(table(predictedClasses))
  
  save(predictedClasses, file = paste0(rdata,"CaSTLe_Annot_output_",paste0(control_data, "_", data_type),"_cellnames.RData"))
  write.csv(predictedClasses, file = paste0(csvs,"CaSTLe_Annot_output_",paste0(control_data, "_", data_type),"_cellnames.csv"))
  print(xtable(table(predictedClasses)))
  
  svg(paste0(results,"R_02_Distr_",paste0(control_data, "_", data_type),".svg"), width = 6, height = 4)
  
  plt <- as.data.frame(table(predictedClasses)) %>% 
    mutate(proc= round(Freq *100 / sum(Freq),2)) %>%
    # mutate(ordered_CT = fct_reorder(predictedClasses, proc)) %>%
    ggplot(aes(x = predictedClasses, y= proc)) + 
    geom_bar(stat="identity") +
    theme_bw(base_size = 18) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
    theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1)) +
    coord_cartesian(ylim = c(0,45)) +
    # Plot labeling
    # labs(title = "Distribution of annotated cell types",
    #      subtitle = paste("HRO ",control_data," after CaSTLe;", data_type)) +
    ylab("Percent (%)") + 
    xlab("Celltype")
  
  print(plt)
  
  dev.off()
  
  print(plt)
  
}

```

```{r dirs}
dir <- "/data/bioinf/projects/data/2019_schROHT/2021_CaSTLeAnnot/"
input <- paste0(dir,"input/")
rdata <- paste0(dir,"RData/")
results <- paste0(dir,"results/")
csvs <- paste0(dir, "csvs/")
```

# Preprocessing
## Generating the SingleCellExperiment data sets
### The Cowan data

```{r read_in_expr_cowan,eval=F, echo=T}
expression_data <- read_tsv(paste0(input, "exprMatrix.tsv"))
expression_data$gene <- gsub(expression_data$gene, pattern = "\\|.*", replacement = "")
rownames(expression_data) <- expression_data$gene
expression_data <- as.data.frame(expression_data[,-1])
```

```{r read_in_meta_cowan}
meta <- read.table(paste0(input,"meta.tsv"),sep = '\t', header = TRUE)
celltypes <- as.character(meta$cell_type)
meta$new_celltypes <- as.factor(ifelse(startsWith(celltypes, "AC_B_"), "AC_B", 
                                  ifelse(startsWith(celltypes, "CdBC_"), "CdBC", 
                                         ifelse(startsWith(celltypes,"ChBC"), "ChBC", 
                                                ifelse(startsWith(celltypes,"MC_"), "MC", 
                                                       ifelse(endsWith(celltypes,"cone"), "cones", 
                                                              ifelse(startsWith(celltypes, "AC_Y"), "AC_Y", celltypes)))))))
exclude_ct <- c("37-","38-","5-" )

cellID_noNum <- meta[!(meta$new_celltypes %in% exclude_ct),"cell_id"]

annotation <- droplevels(meta[meta$cell_id %in% cellID_noNum,"new_celltypes"])

expression_data <- expression_data[,colnames(expression_data) %in% cellID_noNum]

save(annotation, file = paste0(rdata, "01_Compressed_CellTypes.RData"))
```


```{r sce_cowan,eval=F, echo=T}
sceset <- SingleCellExperiment(assays = list(counts = as.matrix(expression_data)), 
                               colData = annotation, rowData = rownames(expression_data))
exprs(sceset) <- log2(calculateCPM(sceset ,use_size_factors = FALSE) + 1)
# use gene names as feature symbols
rownames(sceset) <- rownames(expression_data)
rowData(sceset)$feature_symbol <- rownames(sceset)
# remove features with duplicated names

isSpike(sceset, "ERCC") <- grepl("^ERCC-", rownames(sceset))
sceset <- calculateQCMetrics(sceset, feature_controls = list("ERCC" = isSpike(sceset, "ERCC")))


saveRDS(sceset, file = paste0(rdata,"01_Roska_RefData.rds"))
rm(expression_data,sceset)
```

## Foveal Cowan

```{r}

expression_data <- read_tsv("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/input/Foveal_exprMatrix.tsv")
expression_data$gene <- gsub(expression_data$gene, pattern = "\\|.*", replacement = "")
rownames(expression_data) <- expression_data$gene
expression_data <- as.data.frame(expression_data[,-1])

meta <- read.table("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/input/Foveal_meta.tsv",sep = '\t', header = TRUE)
celltypes <- as.character(meta$cell_type)
meta$new_celltypes <- as.factor(ifelse(startsWith(celltypes, "AC_B_"), "AC_B", 
                                  ifelse(startsWith(celltypes, "CdBC_"), "CdBC", 
                                         ifelse(startsWith(celltypes,"ChBC"), "ChBC", 
                                                ifelse(startsWith(celltypes, "AC_Y"), "AC_Y", 
                                                       ifelse(startsWith(celltypes,"END_"), "END", 
                                                              ifelse(startsWith(celltypes,"MC_"), "MC", 
                                                                     ifelse(startsWith(celltypes,"HC_"), "HC", 
                                                                            ifelse(startsWith(celltypes,"GC_"), "GC", 
                                                                            ifelse(endsWith(celltypes,"cone"), "cones", 
                                                                            ifelse(startsWith(celltypes,"MO_"), "MO", celltypes)))))))))))

# save(new_celltypes, file = paste0(rdata, "01_Compressed_CellTypes_foveal.RData"))

immune_cells <- c("MAST", "NK", "TCell", "MO")
cellID_noImmune <- meta[!(meta$new_celltypes %in% immune_cells),"cell_id"]

annotation <- droplevels(meta[meta$cell_id %in% cellID_noImmune,"new_celltypes"])

expression_data <- expression_data[,colnames(expression_data) %in% cellID_noImmune]

save(annotation, file = paste0(rdata, "01_Compressed_CellTypes_foveal.RData"))


sceset <- SingleCellExperiment(assays = list(counts = as.matrix(expression_data)), 
                               colData = annotation, rowData = rownames(expression_data))
exprs(sceset) <- log2(calculateCPM(sceset ,use_size_factors = FALSE) + 1)
# use gene names as feature symbols
rownames(sceset) <- expression_data$gene
rowData(sceset)$feature_symbol <- rownames(sceset)
# remove features with duplicated names

isSpike(sceset, "ERCC") <- grepl("^ERCC-", rownames(sceset))
sceset <- calculateQCMetrics(sceset, feature_controls = list("ERCC" = isSpike(sceset, "ERCC")))


saveRDS(sceset, file = paste0(rdata,"01_Roska_RefData_Foveal_adult.rds"))
rm(expression_data,sceset)

```

## Peripheral Cowan

```{r}

expression_data <- read_tsv("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/input/Peripheral_exprMatrix.tsv")
expression_data$gene <- gsub(expression_data$gene, pattern = "\\|.*", replacement = "")
rownames(expression_data) <- expression_data$gene
expression_data <- as.data.frame(expression_data[,-1])

meta <- read.table("/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/Roska_data/input/Peripheral_meta.tsv",sep = '\t', header = TRUE)
celltypes <- as.character(meta$cell_type)
new_celltypes <- as.factor(ifelse(startsWith(celltypes, "AC_B_"), "AC_B", 
                                  ifelse(startsWith(celltypes, "CdBC_"), "CdBC", 
                                         ifelse(startsWith(celltypes,"ChBC"), "ChBC", 
                                                ifelse(startsWith(celltypes, "AC_Y"), "AC_Y", 
                                                       ifelse(startsWith(celltypes,"END_"), "END", 
                                                              ifelse(startsWith(celltypes,"MC_"), "MC", 
                                                                     ifelse(startsWith(celltypes,"HC_"), "HC", 
                                                                            ifelse(startsWith(celltypes,"GC_"), "GC", 
                                                                            ifelse(endsWith(celltypes,"cone"), "cones", 
                                                                                   ifelse(startsWith(celltypes,"FB_"), "FB", 
                                                                            ifelse(startsWith(celltypes,"MO_"), "MO", celltypes))))))))))))

immune_cells <- c("MAST", "NK", "TCell", "MO")
cellID_noImmune <- meta[!(meta$new_celltypes %in% immune_cells),"cell_id"]

annotation <- droplevels(meta[meta$cell_id %in% cellID_noImmune,"new_celltypes"])

expression_data <- expression_data[,colnames(expression_data) %in% cellID_noImmune]

save(annotation, file = paste0(rdata, "01_Compressed_CellTypes_peri.RData"))

sceset <- SingleCellExperiment(assays = list(counts = as.matrix(expression_data)), 
                               colData = annotation, rowData = rownames(expression_data))
exprs(sceset) <- log2(calculateCPM(sceset ,use_size_factors = FALSE) + 1)
# use gene names as feature symbols
rownames(sceset) <- expression_data$gene
rowData(sceset)$feature_symbol <- rownames(sceset)
# remove features with duplicated names

isSpike(sceset, "ERCC") <- grepl("^ERCC-", rownames(sceset))
sceset <- calculateQCMetrics(sceset, feature_controls = list("ERCC" = isSpike(sceset, "ERCC")))


saveRDS(sceset, file = paste0(rdata,"01_Roska_RefData_Peri_adult.rds"))
rm(expression_data,sceset)

```


## The HRO Control 2 data set

```{r rean_sce_ctrl2,eval=F, echo=T}
HRO_Control2 <- read.csv('/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/csvs/Count_data_schROHT_Control2.csv',row.names = 1)
HRO_sceset <- SingleCellExperiment(assays = list(counts = as.matrix(t(HRO_Control2))),
                                   rowData = colnames(HRO_Control2))

# in hemberg lab code --> use.size.factors
exprs(HRO_sceset) <- log2(calculateCPM(HRO_sceset,use_size_factors = FALSE) + 1)
# use gene names as feature symbols
rowData(HRO_sceset)$feature_symbol <- rownames(HRO_sceset)
# remove features with duplicated names

isSpike(HRO_sceset, "ERCC") <- grepl("^ERCC-", rownames(HRO_sceset))
HRO_sceset <- calculateQCMetrics(HRO_sceset, feature_controls = list("ERCC" = isSpike(HRO_sceset, "ERCC")))


saveRDS(HRO_sceset, file = paste0(rdata,"01_HRO_TestData.rds"))
rm(HRO_Control2,HRO_sceset)
```

## The HRO Control 3 data set

```{r rean_sce_ctrl3,eval=F, echo=T}
HRO_Control3 <- read.csv('/data/bioinf/projects/data/2019_schROHT/2021_Foveal_Peripheral_R/csvs/Count_data_schROHT_Control3.csv',row.names = 1)
HRO3_sceset <- SingleCellExperiment(assays = list(counts = t(as.matrix(HRO_Control3))),
                                    rowData = colnames(HRO_Control3))

# in hemberg lab code --> use.size.factors
exprs(HRO3_sceset) <- log2(calculateCPM(HRO3_sceset,use_size_factors = FALSE) + 1)
# use gene names as feature symbols
rowData(HRO3_sceset)$feature_symbol <- rownames(HRO3_sceset)
# remove features with duplicated names

isSpike(HRO3_sceset, "ERCC") <- grepl("^ERCC-", rownames(HRO3_sceset))
HRO3_sceset <- calculateQCMetrics(HRO3_sceset, feature_controls = list("ERCC" = isSpike(HRO3_sceset, "ERCC")))


saveRDS(HRO3_sceset, file = paste0(rdata,"01_HRO_Control3_TestData.rds"))

rm(HRO_Control3,HRO3_sceset)
```

Before runing CaSTLe please ensure that the rownames are included in the SCE data sets, otherwise no gene overlap can be calculated

The actual CaSTLe classification is run using individual script on EVE due to the data set size of the Cowan data.

# Post-processing

## Infer Training performance per celltype from Cowan data set


```{r eval_train_2}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control2_run_Organoid.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control2_Organoid.RData"),
                   data_type = "Organoid", 
                   control_data = "Control_2")
```

```{r plot_train_2, fig.width=15, fig.height=10}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control2_run_Foveal_adult.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes_foveal.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control2_Foveal_adult.RData"),
                   data_type = "Foveal", 
                   control_data = "Control_2")

```

```{r translate_train_2}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control2_run_Peri_adult.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes_peri.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control2_Peri_adult.RData"),
                   data_type = "Peri", 
                   control_data = "Control_2")


```



### HRO Control 3

```{r eval_train_2}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control3_run_Organoid.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control3_Organoid.RData"),
                   data_type = "Organoid", 
                   control_data = "Control_3")
```

```{r plot_train_2, fig.width=15, fig.height=10}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control3_run_Foveal_adult.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes_foveal.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control3_Foveal_adult.RData"),
                   data_type = "Foveal", 
                   control_data = "Control_3")

```

```{r translate_train_2}
analyze_CaSTLe_run(eval_data = paste0(rdata, "CaSTLe_Eval_DF_Control3_run_Peri_adult.RData"), 
                   cellname_vec = paste0(rdata, "01_Compressed_CellTypes_peri.RData"), 
                   annot_data = paste0(rdata,"CaSTLe_Annot_output_Control3_Peri_adult.RData"),
                   data_type = "Peri", 
                   control_data = "Control_3")


```